---
title: "A real forecast: electricity demand, against the tidyverts"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{A real forecast: electricity demand, against the tidyverts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

The other vignettes show the API. This one asks a harder question: **is the
foundation model actually any good?**

We forecast Victorian daily electricity demand with TimesFM 2.5 and put it in
the same mable as seasonal naive, ETS, and ARIMA, then score all four on held-out
data with `fabletools::accuracy()`. Nothing here is special-cased for `zukzeit` — the
foundation model is just another column.

Chunks are not evaluated when the vignette is built, because that would download
a 925 MB checkpoint during `R CMD check`. Every output below was produced by
running the code as written.

## Why electricity demand

Time-series foundation models are pretrained on large corpora and then applied
without task-specific training. TimesFM is a decoder-only transformer pretrained
on roughly 100 billion time points, and its authors report zero-shot accuracy
close to supervised models trained per dataset [@das2024timesfm]. Chronos reports
a similar pattern from a different architecture [@ansari2024chronos].

Where that holds up varies a lot by data, and the benchmark literature is
specific about it. GIFT-Eval evaluates across seven domains and a range of
frequencies precisely because performance is not uniform [@aksu2024gifteval].
Energy is one of the domains where these models tend to do well: daily and
hourly load has strong, stable multi-scale seasonality and long history. Reviews
of foundation models in power-system forecasting treat load as a leading
application while being clear that classical methods remain competitive in
places [@zater2026power].

The other half of that literature matters just as much. Foundation models are
*not* uniformly better — statistical baselines remain hard to beat at very high
frequencies and on short histories, and benchmark work repeatedly recommends
scoring against naive and seasonal-naive baselines before trusting a foundation
model [@aksu2024gifteval]. That is exactly what this vignette does, and it is
why `SNAIVE()` is in the comparison rather than a formality.

So this is a favourable case, chosen deliberately. Run the same comparison on
your own series before concluding anything about it.

## The data

`tsibbledata::vic_elec` is half-hourly operational demand for Victoria,
Australia, 2012–2014, from the Monash Time Series Forecasting Archive
[@godahewa2021monash]. We aggregate to daily totals in gigawatt-hours, which
leaves a clean weekly cycle plus a strong annual temperature effect.

```{r}
library(zukzeit)
library(tsibbledata)
library(tsibble)
library(fable)
library(dplyr)

daily <- vic_elec |>
  index_by(Date = as.Date(Time)) |>
  summarise(Demand = sum(Demand) / 1000) |>
  filter(Date >= as.Date("2012-01-01"))

daily
#> # A tsibble: 1,096 x 2 [1D]
#>   Date       Demand
#>   <date>      <dbl>
#> 1 2012-01-01   228.
#> 2 2012-01-02   275.
#> 3 2012-01-03   259.
#> # ℹ 1,093 more rows
```

Hold out the last 28 days:

```{r}
h <- 28
train <- daily |> filter(Date <= max(Date) - h)
# 1,068 training days, 28 held out
```

## Four models, one mable

`TSFM()` is a `fabletools` model definition, so it composes with everything else
in the tidyverts. Note that only three of these four are *fitted*: TimesFM is
zero-shot, so its "training" just binds the history for the key.

```{r}
fits <- model(
  train,
  snaive  = SNAIVE(Demand ~ lag("week")),
  ets     = ETS(Demand),
  arima   = ARIMA(Demand),
  timesfm = TSFM(Demand,
                 model_id = "google/timesfm-2.5-200m-pytorch",
                 device = "cpu")
)

fits
#> # A mable: 1 x 4
#>     snaive           ets                             arima         timesfm
#>    <model>       <model>                           <model>         <model>
#> 1 <SNAIVE> <ETS(M,Ad,M)> <ARIMA(0,0,1)(0,1,1)[7] w/ drift> <TSFM[timesfm]>
```

`fabletools` selected `ETS(M,Ad,M)` and `ARIMA(0,0,1)(0,1,1)[7] w/ drift` on its
own — both picked up the weekly seasonality.

If `arima` comes back as `<NULL model>` with a warning, install `urca`:
`fable::ARIMA()` needs it for unit-root tests, it is only suggested by `fable`,
and `fabletools::model()` degrades a failed fit to a null model rather than
erroring so that one bad column cannot discard the rest of the mable. That is
usually the cause, not the data.

Forecasting and scoring is the ordinary tidyverts workflow:

```{r}
fc <- forecast(fits, h = h)
accuracy(fc, daily) |> arrange(RMSE)
```

| `.model`  |   RMSE |    MAE |  MAPE |  MASE |
|-----------|-------:|-------:|------:|------:|
| `timesfm` |  14.16 |   6.82 |  4.53 | 0.472 |
| `arima`   |  29.18 |  17.41 | 10.92 | 1.203 |
| `ets`     |  31.85 |  19.89 | 12.34 | 1.375 |
| `snaive`  |  33.64 |  23.24 | 13.86 | 1.607 |

Zero-shot TimesFM roughly halves the error of the best fitted classical model on
this window. Read MASE first: it scales error by the in-sample seasonal-naive
benchmark, so **below 1 is better than seasonal naive**. TimesFM is at 0.47;
ARIMA and ETS are both above 1, meaning that on these 28 days they did worse
than the naive benchmark they were fitted to beat.

Two honest caveats. This is a single 28-day window, so treat it as one
observation, not an estimate of expected performance — `fabletools::stretch_tsibble()`
gives proper rolling-origin evaluation, and `vignette("rolling-origin-tuning")`
shows that pattern. And the checkpoint was pretrained on a large corpus whose
exact contents are not public, so contamination cannot be ruled out for a public
benchmark dataset like this one.

## Prediction intervals

The forecast carries predictive quantiles, not just a point:

```{r}
tf <- fc |> filter(.model == "timesfm")
quantile(tf$Demand, 0.1)   #> 221.6 206.6 186.8 ...
quantile(tf$Demand, 0.5)   #> 230.4 214.8 194.0 ...
quantile(tf$Demand, 0.9)   #> 240.9 225.5 204.9 ...

hilo(tf, 60)
#> # A tsibble: 3 x 5 [1D]
#>   .model  Date            Demand .mean                  `60%`
#>   <chr>   <date>          <dist> <dbl>                 <hilo>
#> 1 timesfm 2014-12-04 quantile[3]  231. [223.7762, 238.2957]60
#> 2 timesfm 2014-12-05 quantile[3]  215. [208.6656, 222.8651]60
#> 3 timesfm 2014-12-06 quantile[3]  195. [188.6090, 202.1633]60
```

`hilo(tf, 80)` is the one to watch out for: it returns an `NA` lower bound.
`hilo()` computes `(1 - 80/100)/2`, which in floating point is
`0.09999999999999998` — about two ulps *below* `0.1`. TimesFM's lowest trained
quantile is exactly `0.1`, and `distributional` will not extrapolate past the
outermost percentile it holds, so the lower bound comes back missing. Widening
the request does not help, because the checkpoint emits only `0.1` through `0.9`
and nothing outside. Use an interval that lands inside the trained grid, as
above, or read the quantiles directly.

## Where the point forecast lives

`fabletools` derives `.mean` from the distribution, so for a `TSFM()` fable it is
the distribution's *mean*. The engine's own point forecast is the exactly
evaluated median, which `median()` recovers on any route:

```{r}
median(tf$Demand)
```

The plain-R and `as_fable()` routes report that median in `.mean` directly. The
distributions are identical across routes; only the summary differs.

## Scaling to a panel

`fabletools` evaluates one key at a time, so `TSFM()` runs the engine once per
series. The checkpoint is loaded once and reused from the resident cache, but the
forward passes are not batched. For a wide panel — `aus_retail` has 152 series —
the batched route is much faster:

```{r}
retail_fc <- forecast(
  zuk_pretrained("google/timesfm-2.5-200m-pytorch", device = "cpu"),
  aus_retail, h = 12,
  index = "Month", key = "Series ID", target = "Turnover"
)
nrow(retail_fc)                      #> 1824  (152 series x 12 months)
fabletools::as_fable(retail_fc)
```

All 152 series in one batched call takes about 10 seconds on CPU, including
loading the checkpoint.

Use `TSFM()` when you want the model *inside* a mable next to other models, as
above; use `forecast()` plus `as_fable()` when you want throughput.

## What to take from this

On daily electricity demand — long history, strong seasonality, a domain the
literature flags as favourable — a zero-shot foundation model beat three fitted
classical models by a wide margin, with no tuning and no training.

That result is real but narrow. The same benchmark literature that reports strong
zero-shot performance also reports that these models lose to simple baselines on
short, noisy, or very high-frequency series. `zukzeit` is built so the comparison
above is cheap to run: put `TSFM()` in the mable next to `SNAIVE()` and let
`accuracy()` settle it on your data.

## References
