---
title: "Tuning context length with rolling-origin resampling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tuning context length with rolling-origin resampling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

has_tuning <- requireNamespace("parsnip", quietly = TRUE) &&
  requireNamespace("workflows", quietly = TRUE) &&
  requireNamespace("tune", quietly = TRUE) &&
  requireNamespace("dials", quietly = TRUE) &&
  requireNamespace("rsample", quietly = TRUE) &&
  requireNamespace("yardstick", quietly = TRUE)
knitr::opts_chunk$set(eval = has_tuning)
```

How much history should a foundation model condition on? More context is not
always better — it can dilute recent signal and cost compute. Because `zukzeit`
exposes `context_length` as a tunable `dials` parameter, you can answer this
empirically with the standard tidymodels tuning machinery over **rolling-origin**
(time-aware) resamples, rather than random cross-validation that would leak the
future into the past.

The code below uses the executable `stub` fixture to demonstrate the mechanics,
and every output is produced when this vignette is built. It is not a
foundation-model benchmark: the stub forecasts the last observed value, so its
score cannot depend on how much history it keeps. Repeat the workflow with
TimesFM for a result that means something.

## Some history to search over

```{r}
library(zukzeit)

set.seed(20260904)
months <- seq(as.Date("2015-01-01"), by = "month", length.out = 96)
history <- data.frame(
  store = "north",
  month = months,
  sales = 100 + 0.8 * seq_along(months) +
    10 * sin(seq_along(months) * pi / 6) + rnorm(length(months), sd = 2)
)
head(history, 3)
```

## The spec

Mark `context_length` for tuning; everything else is fixed through the engine:

```{r}
library(parsnip)
library(workflows)
library(tune)
library(dials)
library(rsample)

spec <- zuk_reg(context_length = tune()) |>
  set_engine("zukzeit",
             model_id = "stub",
             index    = "month",
             id       = "store")

wf <- workflow() |>
  add_variables(outcomes = sales, predictors = c(month, store)) |>
  add_model(spec)
```

`add_variables()` matters here. `index` and `id` name columns the engine reads
out of the data it is handed, and `add_formula()` keeps only the terms in the
formula — which drops `store` before the engine sees it, and the fit then fails
on a missing column.

## Rolling-origin resamples

`rsample::sliding_period()` builds analysis/assessment splits that always move
forward in time — the assessment window is strictly after its analysis window:

```{r}
folds <- sliding_period(
  history,
  index       = month,
  period      = "year",
  lookback    = Inf,   # expanding origin: all history up to each split
  assess_stop = 1      # score one year ahead
)
folds
```

## Search

Give `context_length` a range and let `tune_grid()` score each candidate:

```{r, message = FALSE, warning = FALSE}
grid <- grid_regular(context_length(range = c(12L, 60L)), levels = 4)

res <- tune_grid(
  wf,
  resamples = folds,
  grid      = grid,
  metrics   = yardstick::metric_set(yardstick::rmse, yardstick::mae)
)

show_best(res, metric = "rmse")
```

```{r}
best <- select_best(res, metric = "rmse")
best

final_fit <- finalize_workflow(wf, best) |>
  fit(data = history)
final_fit
```

`finalize_workflow()` substitutes the chosen value back into the spec through
the specification's `update()` method, and the refit engine then keeps only that
much history.

## Notes

- `context_length` bounds the most recent history retained by each fitted
  engine; the end-to-end behavior is covered with the stub in package tests.
- The heavy per-candidate cost is inference, not training (these are zero-shot
  models), so keep the grid small and prefer `sliding_period()` windows that
  reflect your real forecast horizon.
- Scores that are identical across the grid, as above, mean the model is not
  actually using the extra history — which for the stub is the correct answer,
  and for a real checkpoint is a finding worth investigating.
