---
title: "Zero-shot forecasting with TimesFM 2.5"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Zero-shot forecasting with TimesFM 2.5}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

TimesFM 2.5 is the first architecture `zukzeit` runs natively: the checkpoint is
loaded into R `torch`, and no Python runs at inference time. This vignette walks
the whole public surface on one panel, from discovery to forecast.

Chunks are not evaluated when the vignette is built — that would download a
925 MB checkpoint during `R CMD check` — but every output below was produced by
running the code as written.

## Find a checkpoint before downloading anything

`zuk_models()` is offline and static. It never fetches a config or a weight
file, so you can inspect what is available, how large it is, and what licence
its weights carry before committing to a download:

```{r}
library(zukzeit)

zuk_models()
#>                          model_id architecture     state max_context  n_params
#> 1 google/timesfm-2.5-200m-pytorch      timesfm supported       16256 231289280
#>   size_bytes cached    license
#> 1  925181104   TRUE Apache-2.0
```

The default lists only checkpoints that have passed both release gates —
architecture conformance and numerical parity against the pinned reference. The
`license` column is the **weight** licence, not the package's.

`cached` is computed with local-only probes, so it tells you whether a download
is pending without touching the network. To fetch the files explicitly:

```{r}
zuk_download("google/timesfm-2.5-200m-pytorch")
```

## Load the model and check what it can do

```{r}
model <- zuk_pretrained("google/timesfm-2.5-200m-pytorch", device = "cpu")

zuk_capabilities(model)
#> <zuk_capabilities>
#>   architecture:      timesfm
#>   max_context:       16256        max_horizon: 1024
#>   quantiles:         native        levels: 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9
#>   multivariate:      FALSE        samples:   FALSE
#>   past_covariates:   FALSE        future_covariates: FALSE
#>   static_covariates: FALSE        fine_tunable: FALSE
#>   license:           Apache-2.0
```

Capabilities are enforced *before* inference, so an unsupported request costs
nothing. The `FALSE` fields are honest: contract v1 has no channel for
covariates or multivariate targets, so `zukzeit` refuses rather than silently
ignoring them.

`max_context` is the best case. TimesFM's 16,384 positions cover the context and
the forecast together, so the history actually used is
`16384 - ceiling(h / 128) * 128` — 16,256 observations at short horizons, 15,360
at the maximum horizon of 1,024. Longer histories are truncated to the most
recent values, per series.

## Forecast a panel

Any data frame with an index, a key, and a numeric target will do; a `tsibble`
works too, with the index and key inferred.

```{r}
set.seed(20260813L)
n <- 120L
panel <- do.call(rbind, lapply(c("north", "south"), function(k) {
  data.frame(
    date   = as.Date("2024-01-01") + seq_len(n) - 1L,
    sales  = 120 + seq_len(n) * 0.25 +
             10 * sin(seq_len(n) * 2 * pi / 7) + rnorm(n, sd = 2),
    region = k
  )
}))

fc <- forecast(
  model, panel,
  h = 14,
  index = "date", key = "region", target = "sales",
  quantile_levels = c(0.1, 0.5, 0.9)
)
fc
#> <zuk_forecast> of "sales"
#> 2 series x horizon, quantile levels 0.1, 0.5, and 0.9
#>    region       date    .mean .distribution
#> 1   north 2024-04-30 162.6610   quantile[3]
#> 2   north 2024-05-01 155.9382   quantile[3]
#> 3   north 2024-05-02 146.4727   quantile[3]
#> 4   north 2024-05-03 142.3131   quantile[3]
#> 5   north 2024-05-04 144.9197   quantile[3]
```

Both series are forecast in **one batched call**. `.mean` is the engine's
exactly evaluated median, and `.distribution` is a `distributional` vector
carrying the requested quantiles.

`as.data.frame()` gives a plain frame with no attributes attached:

```{r}
head(as.data.frame(fc), 3)
#>   region       date    .mean .distribution
#> 1  north 2024-04-30 162.6610   quantile[3]
#> 2  north 2024-05-01 155.9382   quantile[3]
#> 3  north 2024-05-02 146.4727   quantile[3]
```

Requesting a quantile the checkpoint was not trained on fails before any tensor
work, with a classed condition rather than a message you have to parse:

```{r}
forecast(model, panel, h = 14, index = "date", key = "region",
         target = "sales", quantile_levels = c(0.05, 0.5))
#> Error in `forecast()`:
#> ! The checkpoint does not emit every requested quantile level.
#> ✖ Unsupported: 0.05.
#> ℹ Supported levels: 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, and 0.9.
```

The condition carries the leaf class `zuk_error_quantile_levels` and the policy
parent `zuk_error_recoverable`, so a caller can branch on the outcome without
matching message text.

## Into the tidyverts

`fabletools::as_fable()` converts the batched forecast into a fable, so it flows
into `accuracy()`, reconciliation, and the tidyverts plotting stack:

```{r}
fabletools::as_fable(fc)
#> # A fable: 28 x 4 [1D]
#> # Key:     region [2]
#>    region date       .mean       sales
#>    <chr>  <date>     <dbl>      <dist>
#>  1 north  2024-04-30  163. quantile[3]
#>  2 north  2024-05-01  156. quantile[3]
#>  3 north  2024-05-02  146. quantile[3]
#> # ℹ 25 more rows
```

## Composing with `TSFM()`

To compare a foundation model against ARIMA or ETS in one mable, use `TSFM()`
inside `fabletools::model()`:

```{r}
library(tsibble)

fits <- fabletools::model(
  as_tsibble(panel, key = region, index = date),
  timesfm = TSFM(sales, model_id = "google/timesfm-2.5-200m-pytorch",
                 device = "cpu")
)
fits
#> # A mable: 2 x 2
#> # Key:     region [2]
#>   region         timesfm
#>   <chr>          <model>
#> 1 north  <TSFM[timesfm]>
#> 2 south  <TSFM[timesfm]>

fabletools::forecast(fits, h = 14)
#> # A fable: 28 x 5 [1D]
#> # Key:     region, .model [2]
#>    region .model  date             sales .mean
#>    <chr>  <chr>   <date>          <dist> <dbl>
#>  1 north  timesfm 2024-04-30 quantile[3]  163.
#>  2 north  timesfm 2024-05-01 quantile[3]  156.
#>  3 north  timesfm 2024-05-02 quantile[3]  147.
#> # ℹ 25 more rows
```

Two things are worth knowing about this route.

**It is per key, not batched.** `fabletools` trains and forecasts one series at
a time, so the engine sees a batch of one per key. The checkpoint is loaded once
and served from the resident cache for every remaining key, but the forward
passes are not batched. For a wide panel, prefer
`forecast(model, panel, h) |> fabletools::as_fable()`.

**`.mean` is the distribution's mean here**, because `fabletools` derives it
from the distribution. The plain-R and `as_fable()` routes report the engine's
exact median instead. The distributions are identical; only the summary differs.
`median()` recovers the engine's point forecast on any route:

```{r}
median(fabletools::forecast(fits, h = 14)$sales)
```

## Into tidymodels

```{r}
library(parsnip)

spec <- zuk_reg() |>
  set_engine("zukzeit",
             model_id = "google/timesfm-2.5-200m-pytorch",
             device = "cpu", index = "date", id = "region")

fitted <- fit(spec, sales ~ 1, data = panel)
future <- expand.grid(
  date   = max(panel$date) + 1:14,
  region = c("north", "south")
)
predict(fitted, new_data = future)
```

`predict()` returns a tidymodels-conformant frame, row-aligned to `new_data`,
with `.pred`, `.pred_lower`, `.pred_upper`, and one `.pred_qXX` column per
fitted quantile level.

## Managing the checkpoint

A constructed handle is held in a bounded, session-level cache — by default one
model, controlled by `options(zuk.max_loaded_models)`. That is what keeps a
`tune_grid()` or a per-key `TSFM()` run from reloading 925 MB repeatedly.

```{r}
zuk_cache_status()
#>                                model_id                                 revision
#> 1       google/timesfm-2.5-200m-pytorch 1d952420fba87f3c6dee4f240de0f1a0fbc790e3
#> 2 ibm-granite/granite-timeseries-ttm-r2 d6a79570cac0f33d526601cd3a0fc7c80a8f9a2f
#>   device disk_cached resident size_bytes           last_used
#> 1    cpu        TRUE     TRUE  925181104 2026-08-13 21:20:04
#> 2   <NA>       FALSE    FALSE    3240592                <NA>
```

`disk_cached` and `resident` are deliberately separate: the first is about
`hfhub`'s files on disk, the second about a constructed module in this session.
`zuk_unload()` releases resident handles and never deletes downloaded files —
disk lifetime stays with `hfhub`.

```{r}
zuk_unload()
```

## Reproducibility

Support claims are pinned to immutable revisions. The catalogue resolves
checkpoint `1d952420fba87f3c6dee4f240de0f1a0fbc790e3`, never a moving `main`,
and `zuk_pretrained()` rejects any other revision for a curated model. The
committed parity fixtures were generated from official source commit
`3dae50b20d7a724981e8ea36cda75578f80dd2dc`; see
`tests/testthat/fixtures/README.md` for the generator, the tolerance, and how to
regenerate them.
