---
title: "Exercising the forecasting workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exercising the forecasting workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

has_tidyverts <- requireNamespace("tsibble", quietly = TRUE) &&
  requireNamespace("fabletools", quietly = TRUE)
has_tidymodels <- requireNamespace("parsnip", quietly = TRUE) &&
  requireNamespace("workflows", quietly = TRUE)
has_tuning <- has_tidymodels &&
  requireNamespace("tune", quietly = TRUE) &&
  requireNamespace("rsample", quietly = TRUE) &&
  requireNamespace("dials", quietly = TRUE)
```

This vignette exercises the engine and adapter surfaces with `stub`, a
weight-free random-walk fixture. It is not a foundation model — it exists so the
engine and its adapters can be demonstrated and tested without a download. Every
output below is produced when this vignette is built.

For the same workflow driven by a real checkpoint, see
`vignette("timesfm-zero-shot")`.

## Load a model

`tsfm_pretrained()` returns a uniform model handle. The stub branch is entirely
local and deterministic:

```{r}
library(tsfm)

model <- tsfm_pretrained("stub")
```

Before doing any work, ask what the model can do. The capability report is
checked *before* inference, so an unsupported request fails fast with a clear
message instead of deep inside a tensor op:

```{r}
tsfm_capabilities(model)
```

Requests are validated against that report at the engine boundary, not after it:

```{r, error = TRUE}
forecast(model, data.frame(t = 1:10, y = as.numeric(1:10)),
         h = 5, index = "t", target = "y", quantile_levels = c(0.1, 1.5))
```

## A panel to forecast

Two monthly series, split into observed history and the twelve months held back
to score against:

```{r}
set.seed(1)
months <- seq(as.Date("2015-01-01"), by = "month", length.out = 96)
panel <- data.frame(
  store = rep(c("north", "south"), each = length(months)),
  month = rep(months, 2),
  sales = c(
    100 + 0.8 * seq_along(months) + 10 * sin(seq_along(months) * pi / 6),
    50 + 0.3 * seq_along(months) + 4 * cos(seq_along(months) * pi / 6)
  ) + rnorm(2 * length(months), sd = 2)
)

split_at <- max(months) - 365
history <- panel[panel$month <= split_at, ]
actuals <- panel[panel$month > split_at, ]
nrow(history)
nrow(actuals)
```

## Path 1 — the `forecast()` convenience surface

Hand a panel of history and a horizon; get a forecast object back. This path has
no optional dependencies at all — a plain data frame is enough, as long as the
index and target columns are named:

```{r}
fc <- forecast(model, history, h = 12,
               index = "month", key = "store", target = "sales",
               quantile_levels = c(0.1, 0.5, 0.9))

fc
```

The forecast carries both views of the same predicted quantiles: a
`distributional` column for the tidyverts, and the raw quantile matrix for the
tidymodels prediction columns.

```{r}
head(as.data.frame(fc), 3)
```

Note that the future index comes back as `Date`, the type it went in as. The
engine extends an index in its own units, so a `yearmonth` panel returns
`yearmonth`, an hourly `POSIXct` panel returns `POSIXct`, and the result joins
straight back onto the data it came from.

## Path 2 — the tidyverts

`tsfm` registers a method on `fabletools::as_fable()` rather than exporting a
generic of its own, so the verb keeps working when fabletools and `tsfm` are
both attached. Call it qualified, or `library(fabletools)` first.

```{r, eval = has_tidyverts}
# yearmonth is the natural index for monthly data: calendar months are not
# equally spaced in days, so a Date index would put the forecasts a few days
# off the observed month starts and nothing would join.
as_monthly <- function(data) {
  data$month <- tsibble::yearmonth(data$month)
  tsibble::as_tsibble(data, index = month, key = store)
}
history_tsbl <- as_monthly(history)
actuals_tsbl <- as_monthly(actuals)

fabletools::as_fable(forecast(model, history_tsbl, h = 12)) |>
  fabletools::accuracy(actuals_tsbl)
```

`TSFM()` is the other tidyverts route: it composes inside `fabletools::model()`
next to any other model definition, one key at a time.

```{r, eval = has_tidyverts}
fits <- fabletools::model(history_tsbl, tsfm = TSFM(sales, model_id = "stub"))
fits

fabletools::forecast(fits, h = 3)
```

## Path 3 — a tidymodels workflow

The tidymodels surface is a `parsnip` spec inside a `workflow()`. The model is
chosen by the engine's `model_id`; **everything else stays the same.**

```{r, eval = has_tidymodels}
library(parsnip)
library(workflows)

spec <- tsfm_reg(context_length = 512L) |>
  set_engine("tsfm",
             model_id = "stub",
             index    = "month",
             id       = "store")

wf <- workflow() |>
  add_variables(outcomes = sales, predictors = c(month, store)) |>
  add_model(spec)

fitted <- fit(wf, data = history)
head(predict(fitted, new_data = actuals[, c("store", "month")]), 3)
```

`add_variables()` rather than `add_formula()` is deliberate. `index` and `id`
name columns the engine reads out of the data it is handed, and a formula
preprocessor keeps only the terms in the formula — which would drop `store`
before the engine ever sees it.

## Tuning the context length

`context_length` is a tunable `dials` parameter, so a rolling-origin search over
`rsample` resamples works out of the box:

```{r, eval = has_tuning, message = FALSE, warning = FALSE}
library(tune)
library(rsample)

tuned <- tune_grid(
  workflow() |>
    add_variables(outcomes = sales, predictors = c(month, store)) |>
    add_model(tsfm_reg(context_length = tune()) |>
                set_engine("tsfm", model_id = "stub",
                           index = "month", id = "store")),
  resamples = sliding_period(history[history$store == "north", ],
                             index = month, period = "year",
                             lookback = Inf, assess_stop = 1),
  grid = dials::grid_regular(context_length(range = c(12L, 48L)), levels = 3)
)

show_best(tuned, metric = "rmse")
```

Every candidate scores the same here, and that is the expected answer: the stub
forecasts the last observed value, so how much history it keeps cannot change
its point forecast. Run the same search against TimesFM for a result that means
something — `vignette("rolling-origin-tuning")` walks it in more detail.

## Clean up

The constructed handle is held in a bounded in-session cache. Evicting it does
not touch anything on disk:

```{r}
tsfm_cache_status()

tsfm_unload("stub")
```

## Status

The public workflow above is real, but its model is deliberately only a test
fixture. TimesFM 2.5 is the first implemented release model; see
`vignette("timesfm-zero-shot")` for the same surfaces driven by real weights.
TTM remains a registered scaffold, and Chronos-2 is rejected before any network
or tensor work until its native port lands.
