---
title: "Exercising the forecasting workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exercising the forecasting workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette exercises the engine and adapter surfaces with `stub`, a
weight-free random-walk fixture. It is not a foundation model — it exists so the
engine and its adapters can be demonstrated and tested without a download.

For the same workflow driven by a real checkpoint, see
`vignette("timesfm-zero-shot")`.

## Load a model

`tsfm_pretrained()` returns a uniform model handle. The stub branch is entirely
local and deterministic:

```{r}
library(tsfm)

model <- tsfm_pretrained("stub")
```

Before doing any work, ask what the model can do. The capability report is
checked *before* inference, so an unsupported request fails fast with a clear
message instead of deep inside a tensor op:

```{r}
tsfm_capabilities(model)
#> <tsfm_capabilities>
#>   architecture:      stub
#>   max_context:       512         quantiles: native
#>   multivariate:      FALSE       samples:   FALSE
#>   past_covariates:   FALSE       future_covariates: FALSE
#>   static_covariates: FALSE       fine_tunable: FALSE
#>   license:           MIT
```

## Path 1 — the `forecast()` convenience surface (tsibble)

Hand a panel of history and a horizon; get a forecast object back that converts
to a `fable` for the tidyverts accuracy/reconciliation stack:

```{r}
library(tsibble)

fc <- forecast(model, tourism_history, h = 12,
               quantile_levels = c(0.1, 0.5, 0.9))

fc |>
  fabletools::as_fable() |>
  fabletools::accuracy(tourism_actuals)
```

`tsfm` registers a method on `fabletools::as_fable()` rather than exporting a
generic of its own, so the verb keeps working when fabletools and `tsfm` are
both attached. Call it qualified, or `library(fabletools)` first.

## Path 2 — a tidymodels workflow

The tidymodels surface is a `parsnip` spec inside a `workflow()`. The model is
chosen by the engine's `model_id`; **everything else stays the same.**

```{r}
library(parsnip)
library(workflows)

spec <- tsfm_reg(context_length = 512L) |>
  set_engine("tsfm",
             model_id = "stub",
             index    = "date",
             id       = "store")

wf <- workflow() |>
  add_formula(demand ~ date) |>
  add_model(spec)

fit <- fit(wf, data = train_df)
predict(fit, new_data = future_df)   # tibble: .pred
```

## Tuning the context length

`context_length` is a tunable `dials` parameter, so a rolling-origin search over
`rsample` resamples works out of the box:

```{r}
library(tune)
library(rsample)

spec <- tsfm_reg(context_length = tune()) |>
  set_engine("tsfm", model_id = "stub", index = "date", id = "store")

tune_grid(
  workflow() |> add_formula(demand ~ date) |> add_model(spec),
  resamples = sliding_period(train_df, date, "month"),
  grid      = 5
)
```

## Status

The public workflow above is real, but its model is deliberately only a test
fixture. TimesFM 2.5 is the first implemented release model. Toto 2.0 4M and a
new native Chronos-2 port are required before `0.1.0`; the old Chronos-2 Brulee
bridge remains unregistered reference work only. TTM is not a current target
because the selected small-model slot is probabilistic Toto rather than a
point-only checkpoint.
