---
title: "Using the models"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This article walks the three supported checkpoints end to end: what they are,
how to forecast with them, how capabilities decide what you may ask, and what
Chronos-2 can do that the others cannot.

Chunks are not evaluated when the site is built, because doing so would
download about 1.4 GB of weights and needs the `torch` runtime. Every output
below was produced by running the code exactly as written.

## What is available

`zuk_models()` is offline and static. It never fetches a config or a weight
file, so you can compare cost and capability before committing to a download.

```{r}
library(zukzeit)

zuk_models()[, c("model_id", "architecture", "max_context", "n_params",
                 "size_bytes", "cached")]
#>                          model_id architecture max_context  n_params
#> 1 google/timesfm-2.5-200m-pytorch      timesfm       16256 231289280
#> 2                amazon/chronos-2     chronos2        8192 119477664
#> 3             Datadog/Toto-2.0-4m        toto2        4096   4144456
#>   size_bytes cached
#> 1  925181104   TRUE
#> 2  477930472   TRUE
#> 3   16582848   TRUE
```

The three are complementary rather than ranked. TimesFM has by far the longest
context. Toto is 56x smaller than TimesFM and the natural default for CPU work.
Chronos-2 is the only one that accepts covariates:

```{r}
zuk_models()[, c("architecture", "multivariate", "past_covariates",
                 "future_covariates")]
#>   architecture multivariate past_covariates future_covariates
#> 1      timesfm        FALSE           FALSE             FALSE
#> 2     chronos2         TRUE            TRUE              TRUE
#> 3        toto2        FALSE           FALSE             FALSE
```

Those flags describe executable behaviour, not intent. A `TRUE` means there is
a code path with fixtures behind it.

## A forecast

Hand `forecast()` a data frame and name the index and target columns. Nothing
optional is required for this route --- no `tsibble`, no `tidymodels`.

```{r}
set.seed(20260905)
n <- 240
t <- seq_len(n)
series <- data.frame(
  day = seq(as.Date("2024-01-01"), by = "day", length.out = n),
  demand = 500 + 1.2 * t + 60 * sin(t * 2 * pi / 7) + rnorm(n, 0, 12)
)

model <- zuk_pretrained("Datadog/Toto-2.0-4m")
fc <- forecast(model, series, h = 14, index = "day", target = "demand",
               quantile_levels = c(0.1, 0.5, 0.9))
head(as.data.frame(fc)[, c("day", ".mean")], 4)
#>          day    .mean
#> 1 2024-08-28 800.9774
#> 2 2024-08-29 760.7402
#> 3 2024-08-30 732.6268
#> 4 2024-08-31 747.5360
```

The future index comes back as `Date`, the type it went in as. An index is
extended in its own units, so a `yearmonth` panel returns `yearmonth` and an
hourly `POSIXct` panel returns `POSIXct` --- the result joins straight back
onto the data it came from.

Swapping the checkpoint changes nothing else. On the same series:

```{r}
#  Toto 2.0 4M      800.98  760.74  732.63  747.54
#  Chronos-2        815.58  769.38  734.68  746.35
#  TimesFM 2.5      805.94  765.69  733.85  750.59
```

## Capabilities decide what you may ask

Every checkpoint carries a capability record, and the engine checks a request
against it *before* any download or tensor work.

```{r}
zuk_capabilities(zuk_pretrained("amazon/chronos-2"))
#> <zuk_capabilities>
#>   architecture:      chronos2
#>   max_context:       8192        max_horizon: 1024
#>   quantiles:         native        levels: 0.01, 0.05, 0.1, ..., 0.95, 0.99
#>   multivariate:      TRUE        samples:   FALSE
#>   past_covariates:   TRUE        future_covariates: TRUE
#>   static_covariates: FALSE        fine_tunable: FALSE
#>   license:           Apache-2.0
```

Chronos-2 emits 21 trained quantile levels; TimesFM and Toto emit nine. Asking
for a level a checkpoint was not trained on is refused, with the supported set
in the message:

```{r}
forecast(model, series, h = 14, index = "day", target = "demand",
         quantile_levels = c(0.1, 0.123, 0.9))
#> Error:
#> ! The checkpoint does not emit every requested quantile level.
#> ✖ Unsupported: 0.123.
#> ℹ Supported levels: 0.01, 0.05, 0.1, 0.15, 0.2, ..., 0.95, and 0.99.
```

Failures are classed, so a caller can branch on them without matching message
text. See `` ?`zuk-conditions` ``.

## Covariates

Chronos-2 accepts series that inform a forecast without being forecast
themselves. Name them with `covariates`, and supply their future values in a
separate frame when you know them.

Here demand carries an irregular promotion effect --- irregular on purpose,
because nothing in the target's own seasonality predicts it:

```{r}
set.seed(11)
promo <- as.numeric(runif(n) < 0.15)
panel <- data.frame(
  store = "flagship",
  day = series$day,
  demand = 500 + 1.2 * t + 40 * sin(t * 2 * pi / 7) + 150 * promo + rnorm(n, 0, 8),
  promo = promo
)

future_days <- seq(max(panel$day) + 1, by = "day", length.out = 10)
planned <- data.frame(store = "flagship", day = future_days,
                      promo = c(0, 0, 1, 1, 0, 0, 0, 1, 0, 0))

model <- zuk_pretrained("amazon/chronos-2")
informed <- forecast(model, panel, h = 10, index = "day", key = "store",
                     target = "demand", covariates = "promo", future = planned)
blind <- forecast(model, panel, h = 10, index = "day", key = "store",
                  target = "demand")
```

```{r}
#>         day promo informed blind   lift
#>  2024-08-28     0    796.3 810.7  -14.4
#>  2024-08-29     0    773.3 773.2    0.1
#>  2024-08-30     1    854.8 752.8  102.0
#>  2024-08-31     1    887.4 762.5  124.9
#>  2024-09-01     0    796.4 795.1    1.3
#>  2024-09-02     0    834.5 819.1   15.3
#>  2024-09-03     0    830.6 829.4    1.3
#>  2024-09-04     1    950.5 814.1  136.4
#>  2024-09-05     0    769.7 781.0  -11.3
#>  2024-09-06     0    766.5 761.2    5.3
#>
#> mean lift on planned promo days : 121.1
#> mean lift on ordinary days      : -0.4
#> true promo effect in history    : 117.2
```

The model recovers a promotion effect of 121 against a true effect of 117, and
leaves ordinary days alone.

A covariate earns its keep when it carries information the target's own history
does not. Running the same example with a strictly *periodic* promotion instead
moves the forecast by only a point or two, because a fourteen-day cycle is
already visible in the target and the model has been reading it all along.

A covariate whose future you do not know is still useful --- omit it from
`future` and it is treated as past-only. The distinction is inferred, so there
is nothing extra to declare.

Asking a checkpoint that cannot do this is refused, and says why:

```{r}
forecast(zuk_pretrained("Datadog/Toto-2.0-4m"), panel, h = 10, index = "day",
         key = "store", target = "demand", covariates = "promo",
         future = planned)
#> Error:
#> ! This architecture does not accept grouped inputs.
#> ℹ It targets contract 1.0.0; grouped inputs need 1.1.0.
```

## Cross-series learning

By default each series in a panel is forecast on its own, so a result depends
on that series and the checkpoint, not on what else shared the call. Naming a
column with `group` opts in to letting series inform one another:

```{r}
forecast(model, panel, h = 10, index = "day", key = "store",
         target = "demand", group = "region")
```

Series sharing a `region` value then form one task. This needs a checkpoint
declaring `multivariate`, since a task with several series has several targets.

## The tidyverts

`TSFM()` composes inside `fabletools::model()` next to any other model
definition, and regressors use fable's own vocabulary:

```{r}
library(fabletools)

fits <- model(
  history_tsbl,
  chronos = TSFM(demand ~ xreg(promo), model_id = "amazon/chronos-2"),
  arima = fable::ARIMA(demand)
)
forecast(fits, new_data = future_tsbl)
```

`xreg()` marks a regressor whose future values arrive in `new_data`;
`past_xreg()` marks one with no future. Note that `fabletools` evaluates one
key at a time, so this route gives covariates but not cross-series learning ---
for that, use the batched `forecast()` route above.

## Lifecycle

Checkpoints are downloaded once and reused. `zuk_download()` prefetches without
constructing anything, `zuk_cache_status()` separates what is on disk from what
is resident in the session, and `zuk_unload()` frees the latter without
deleting the former.

```{r}
zuk_download("Datadog/Toto-2.0-4m")
zuk_cache_status()
zuk_unload()
```

Handle reuse is bounded by `options(zuk.max_loaded_models)`, so forecasting a
wide panel loads a checkpoint once rather than once per series.
